---
title: "Updating a Theory"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{updating_theory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(theorytools)
```

FAIR theories are designed to be living documents - open to updates, refinements, and community contributions.
In this vignette, we describe how to update a FAIR theory by creating a local copy, editing it, and sharing the updated theory either as a new theory with metadata linking to its predecessor, or as a contribution to the original project.

## Step 1: Accessing an Existing Theory

We describe two cases: one where the theory is archived only on Zenodo, and one where the theory is also archived on GitHub.

### A Theory on Zenodo

If the theory only exists on Zenodo,
then we can download a local copy by running the code below,
which is demonstrated here with Morris' Tripartite Model (see [this vignette](https://cjvanlissa.github.io/theorytools/articles/causal-inference.html)).
Change the `path` to a directory on your hard drive where you want to store the theory.

```{r eval = FALSE}
download_theory(
  id = "https://doi.org/10.5281/zenodo.14921521",
  path = "c:/theories/tripartite_downloaded")
```

### A Theory on GitHub

Navigate to the GitHub repository of the FAIR theory you want to update.
Click the **Fork** button in the upper right corner to create your own copy of the repository under your GitHub account (for more information, see [the GitHub documentation](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo)).

This gives you a space to experiment and propose changes without affecting the original repository.

Once you've forked the repository, clone it to your local machine by running the following code in R ([see here for more information](https://docs.github.com/en/repositories/creating-and-managing-repositories/cloning-a-repository)):


```{r eval = FALSE}
download_theory(
  id = "https://github.com/cjvanlissa/tripartite_model.git",
  path = "c:/theories/tripartite_clone")
```

## Step 2: Make Changes to the Theory

You can now:

* Edit the theory implementation (e.g., `tripartite_model.txt`, or other files in which the theory is implemented, see the README if you're not sure which files these are)
* Modify or expand the README
* Update metadata or ontology elements
* Add documentation or related materials
* Increment the version number of the theory

Track your changes using Git, and push these changes to your GitHub repository by running:

```{r eval = FALSE}
worcs::git_update("Describe your changes to the theory.")
```


## Step 3 (Optional): Create a Pull Request

If you'd like the original authors to consider incorporating your updates into the main project:

1. Go to your forked repository on GitHub.
2. Click **"Compare & pull request"** (see [here for more information](https://github.blog/developer-skills/github/beginners-guide-to-github-creating-a-pull-request/))
3. Write a short message explaining your changes and their motivation.
4. Submit the pull request.

The maintainers of the original theory can then review your suggestions and decide whether to merge them.

## Step 4 (Optional): Create an Independent FAIR Theory

Sometimes theories branch off and go their separate way;
if this is the case for your theory, you should follow the steps in [this vignette](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html)) to archive your FAIR theory on Zenodo.

Importantly, if your FAIR theory originated from a previous FAIR theory, you should document its predecessor in the metadata.

To do so, [edit the theory's metadata](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html#entering-meta-data). Under the `Related works` field, add the DOI of the theory's predecessor. In the `Relation` field, select `Is derived from`.
